Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-13 demystify.tex
updates

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
\section{Why the US and not other countries?}
TODO

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\subsection{Self-written Letters are Bad} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \textbf{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

Most well-known profs. would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). However, many international students do not know this (e.g., normal at their university) and might be in a situation they do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer or relatively unknown in the field.  In either case, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

\subsubsection{Letters from Admin people}  
The ``department head'', ``dean'' or ``director'' at your university might not be well-known to adcom members (sorry, we likely recognized researchers, not admin people), so it's fine to to have "less well-known" people write for you as long as they can write a good letter about your research ability).

@@

\begin{commentbox}[Vu:]
  I often read SOP and LORs first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something stand out in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\subsection{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicitly say something about them in your statement.

@@

@@

\section{Standard Tests (GRE, IELTS, etc), Work Experiences, Online Courses, and Certificates}\label{sec:standard-tests}

@@

\subsection{Work Experience} Work experience, especially if they are \emph{not} research-based, \emph{do not} carry much weight for PhD (they can help for MS application though). Similarly, LoR from your supervisor for non-research experience will not count much.

Do not spend much time talking about non-research-related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experience}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LOR (\autoref{sec:lor}).

@@

For each application, the Mason application system compiles a single PDF file, which consists of a summary, transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{sec:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

I read carefully \emph{strong LoRs} and skim weaker ones, noting either strong letters from well-known professors/researchers or generic content. For the \emph{CV}, I look for publications, research experiences, and notable achievements, noting things like ``published papers in top venues'' or ``gold medal in an international competition.''
I read strong \emph{SOP} thoroughly (and skim weak ones). I note whether the SOP is exciting, research-oriented, and tailored to our program. All these notes are entered into the evaluation system.

Finally, I make a decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship. The adcom chair then reviews all evaluations, looks at entered notes, and asks reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject). Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others.

@@

In most cases adcom members, even those reviewing the same application, make decisions independently and do not talk to each other. It's not a strict rule but a common practice to avoid bias and ensure fairness (and also we don't have time to talk to each other).

>>>>

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
% \section{Why the US and not other countries?}
% TODO

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head or dean). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:admin-letters} and \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\subsection{Self-written Letters are Bad} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot  (because the student often has little experience in writing LoRs) and  \red{red flag} such letters.

Most well-known and well-respected profs. would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). However, many international students do not know this (e.g., it is a common practice at their university) and might be in a situation they do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer, relatively unknown in the field, or simply do not know/care enough about you.  In any of these cases, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

\subsection{Letters from Admin people}\label{sec:admin-letters}

Many international students try to get letters from high-ranking admin people in their universities, e.g., department head, dean, or director.  These people, while well-known and respected in your university, might not be too active in research (e.g., they focus on administration) and thus are not recognized by adcom members, who often recognize good researchers (even if they are not in their field'', but not admin people. So it's fine to to have `less well-known'' people write for you as long as they can write a good letter about your research ability.

\begin{commentbox}[Vu:]
  In my experience in reviewing applications, letters from admin people are often generic and do not provide much value.  If you are in this situation, you should find someone else to write for you.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  I often read SOP and LORs first (\autoref{sec:ievaluate}). If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something stand out in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\subsection{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work at such places, then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LORs (\autoref{sec:lor}).

@@

On the other hand, if non-research work (e.g., software development) \emph{do not} carry much weight for PhD (they can help for MS application though). Similarly, LoR from your supervisor for non-research experience will not count much.
So do not spend much time talking about non-research-related job in your SOP.

@@

\section{Standard Tests (GRE, IELTS, etc), Online Courses, and Certificates}\label{sec:standard-tests}

@@

@@

For each application, our system compiles a single PDF file, which consists of a summary (degrees, GPAs, etc), transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually read in this order. I start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{sec:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

I read carefully \emph{strong LoRs} and skim weaker ones, noting either strong letters from well-known professors/researchers or generic content. I skim over \emph{CV} and look for publications, research experiences, and notable achievements, noting things like ``published papers in top venues'' or ``gold medal in an international competition.''
I read strong \emph{SOPs} thoroughly (and skim weak ones). I note whether the SOP is exciting, research-oriented, standing out, e.g., applicant is minority (\autoref{sec:urm}), and tailored to our program (e.g., if they are familiar with the work of some faculty or have talked to them). These notes are entered into the evaluation system.

Finally, I enter my decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship. Note that while the system has other decision options, eg., admit without funding, provisional admission (e.g., if they need to take some courses), I rarely use them, simply because we already have too many good students and do not have the resources to support them.

\subsection{How Decisions are Made}\label{sec:how-decisions}

After all reviewers have evaluated an application, adcom chair will review all evaluations, look at entered notes, and ask reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject).  Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others. In most cases adcom members, even those reviewing the same application, make decisions independently and do not talk to each other (just a common practice to avoid biasing). In some rare cases we might (\autoref{sec:adcom-discuss}).

Even if \emph{all} reviewers recommend acceptance, the application is not automatically accepted, especially if no faculty is willing to advise the student. For example, if the student is interested in a research area that no faculty is working on, or an area where no faculty is taking new students (e.g., AI/ML where faculty likely already has many students) then the student will not be admitted.  This is increasingly common as the number of applicants increases significantly in recent years in CS.

However, if the student has contacted a faculty member and that faculty member is interested in the student and has made this known to the admission committee, then the student is likely to be admitted, \emph{even if the faculty does not review the application or has funding} (thus the importance of contacting faculty). 
Similarly, if the student mentioned a faculty member in their SOP, adcom might ask that faculty member to look at their application and if they are interested in the student.  Even if the student has a weak profile (but still passes the minimum requirement from the university), they might be admitted if a faculty member is willing to take them. Adcom members, especially in the US, are very reluctant to go against the faculty's decision (e.g., if a faculty wants to admit a student, we are not going to reject them).

@@
